---
title: "R: Basic Operations"
output: html_notebook
---

# R: Basic Operations

## Comments

Comments in R are denoted by `#`

```{r}
# This is a comment
```

There is no built-in way to do multi-line comments.

## Mathematical Operations

Mathematical operations in R are handled using the standard 4 symbols `+, -, /, and *` . Operations are calculated using PEMDAS.

### Addition

```{r}
# Addition
4+5
```

### Subtraction

```{r}
# Subtraction
71-43
```

### Multiplication

```{r}
# Multiplication
2*3

```

### Division

```{r}
# Division
80/4
```

## Variable Assignment

Variables in R are traditionally assigned using an arrow sign `<-` , made using a less than carat and a dash. It is valid to use `=` , but not recommended.

```{r}
x <- y <- z <- 10

```

Variable names can not include and spaces or symbols other than the underscore `_` . They cannot begin with numbers, but can include numbers after the first character.

```{r}
v <- c(10, 20, 30)
names(v) <- c("James", "Ola", "maria")
print(v)


```

## Data Types

R has the following data types: numeric, integer, character, logical, and complex.

### Numeric

Decimal values in R are numerics. Numerics are the default data type for numbers.

```{r}

numeric_vector <- c(1, 2.5, 3.14)


is_numeric <- is.numeric(numeric_vector)

print(numeric_vector)
print(is_numeric)

result <- sum(numeric_vector)
print(result)


```

```{r}

```

Note: Numerics are not integers. You can confirm this with both a and b above using the `is.integer()` function.

```{r}
my_integer <- 42L


```

```{r}

```

### Integer

To create an integer in R we can use the `as.integer()` function.

```{r}
number <- 3.14
integer_number <- as.integer(number)
print(integer_number)

character <- "5"
integer_character <- as.integer(character)
print(integer_character)



```

```{r}

```

For shorthand, we can also define an integer by writing the desired integer value, followed by a capital L.

```{r}
my_integer <- 42L

```

```{r}

```

Integers must be whole numbers, if passed a floating value, the number will be rounded to an integer.

```{r}
sample(5.6, 3) 

integer(3.8)


```

This also works with a string.

```{r}

integer1 <- 123
integer2 <- 456

string1 <- as.character(integer1)
string2 <- as.character(integer2)

concatenatedString <- paste(string1, string2) # With space
concatenatedString0 <- paste0(string1, string2) # Without space

print(concatenatedString)
print(concatenatedString0)

concatenatedInteger <- as.integer(concatenatedString0)

print(concatenatedInteger)

```

We would use an integer data-type when we want to represent a whole integer, and don't want to accept any other variations of numbers. The numeric data-type is less strict, it allows us to switch between different number types like integers and floats dynamically.

### Character

A **character** type is used to represent strings in R. Characters can be denoted by writing within quotes.

```{r}

string1 <- "Hello, World!"
string2 <- 'This is a string in R.'


print(string1)
print(string2)


combined_string <- paste(string1, string2)
print(combined_string)


```

#### Concatenating character objects

Character objects can be concatenated into a single variable using the `paste` function

```{r}

string1 <- "Hello"
string2 <- "World"


concatenatedString <- paste(string1, string2)


print(concatenatedString)


```

#### Converting into character objects

We can convert other objects into character objects using the `as.character()` function

```{r}
num_value <- 1234
char_value_from_num <- as.character(num_value)
print(char_value_from_num)


factor_value <- factor(c("apple", "banana", "cherry"))
char_value_from_factor <- as.character(factor_value)
print(char_value_from_factor)


date_value <- as.Date("2024-01-20")
char_value_from_date <- as.character(date_value)
print(char_value_from_date)


```

```{r}
# Note that class returns the data type of our object

```

#### Printing character types

You can print out hardcoded character objects with the `print` function

```{r}

my_string <- "Hello, R!"


print(my_string)


```

You can print out formatted character objects with the `sprintf` function. You can use the following symbols to represent formatted parts of the code, each should be preceded by a `%` symbol:

-   d and i - Integer values

-   s - Character string

-   o - Octal

-   x and X - Hexadecimal

-   f - Double precision value (For fixed point representation, for example: 21.234)

-   e and E - Double precision value (For exponential representation, for example: 2.1234e+01)

```{r}

integer_value <- 123
sprintf("Integer (d): %d", integer_value)
sprintf("Integer (i): %i", integer_value)


string_value <- "Hello, R!"
sprintf("String (s): %s", string_value)


sprintf("Octal (o): %o", integer_value)


sprintf("Hexadecimal lowercase (x): %x", integer_value)
sprintf("Hexadecimal uppercase (X): %X", integer_value)


double_value <- 21.234
sprintf("Double precision (fixed point, f): %.3f", double_value)


sprintf("Double precision (exponential lowercase, e): %.2e", double_value)
sprintf("Double precision (exponential uppercase, E): %.2E", double_value)


```

```{r}

```

#### Return a substring

You can get a substring from a character object using the `substr` function

```{r}

text <- "Hello, R Studio!"


substring <- substr(text, 8, 9)


print(substring)

```

#### Replacing characters in a string

You can replace characters in a string using the `sub` function

```{r}

original_string <- "Hello world, world is beautiful"
new_string_sub <- sub("world", "earth", original_string)
print(new_string_sub)

new_string_gsub <- gsub("world", "earth", original_string)
print(new_string_gsub)

```

### Logical

A logical value is commonly referred to as a boolean value. It is the result of a comparison between two variables. It is represented as `TRUE` or `FALSE.`

```{r}
x <- 5
y <- 10


is_x_greater_than_y <- x > y  
is_x_less_than_y <- x < y     

print(is_x_greater_than_y)  
print(is_x_less_than_y)     


```

```{r}

```

#### Logical Operators

The standard logical operations are:

-   `&` - And

-   `|` - Or

-   `!` - Not

```{r}
a <- TRUE
b <- FALSE
result <- a & b 

```

```{r}
a <- TRUE
b <- FALSE
result <- a | b 

```

```{r}
a <- TRUE
result <- !a 


```

```{r}
x <- 10
y <- 20

if (x > 5 & y > 15) {
  print("Both conditions are true")
}


if (x > 15 | y > 15) {
  print("At least one condition is true")
}

if (!FALSE) {
  print("Not False is True")
}


```

### Complex

A complex object in R is created using the symbol `i` . The use of the i symbol is in reference to imaginary numbers.

```{r}

z <- 3 + 4i


print(z)


real_part <- Re(z)
print(real_part)


imaginary_part <- Im(z)
print(imaginary_part)


modulus <- Mod(z)
print(modulus)


argument <- Arg(z)
print(argument)


conjugate <- Conj(z)
print(conjugate)

```

```{r}

```

## Comparison Operators

You can use the following operators to compare two variables and return a logical object:

-   `<` Less than

-   `>` Greater than

-   `<=` Less than or equal to

-   `>=` Greater than or equal to

-   `==` Is equal to

-   `!=` Is NOT equal to

```{r}

a <- 5
b <- 10

print(a < b)
```

```{r}

print(a > b)  
```

```{r}

print(a <= 5) 

```

```{r}

print(b >= 10) 
```

```{r}

print(a == 5)  
```

```{r}

print(a != b)  
```

## Homework

1.  Create two strings with your first and last name, respectively. Then concatenate them, and print the new value as "My name is \<your newly created concatenated variable will go here\>"

    ```{r}

    firstName <- "Olakunle"
    lastName <- "Kuforiji"


    fullName <- paste(firstName, lastName)


    print(paste("My name is", fullName))

    ```

2.  What values can a BOOLEAN contain?

```{=html}
<!-- -->
```
1.  [**`TRUE`**:]{.underline} This represents the logical value true. It's often the result of logical comparisons or conditions that are met.

2.  [**`FALSE`**]{.underline}: This represents the logical value false. It's the result of logical comparisons or conditions that are not met.

3.  [**`NA`**:]{.underline} This stands for 'Not Available' or 'Missing Value'. It's used in R to represent an unknown or undefined value. In logical operations, if any operand is **`NA`**, the result is often **`NA`**, unless the outcome is certain regardless of the missing value.
